\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-05-07 analysis/paper/paper.tex

From the perspective of clinicians and their patients, the most relevant
quantity is often the 5- or 10-year risk of having a certain event given
the patient's particular circumstances, and not the hazard ratio between
a treatment and control group. Therefore, to make sound clinical
decisions, it is important to accurately estimate the \emph{full} hazard
function, which can then be used to estimate the cumulative incidence
function (CIF). From this perspective, the CIF and the survival function
estimates are now smooth functions of time.

@@

In this article, we present an \proglang{R} package that combines the
ideas of Hanley \& Miettinen into a simple interface. The purpose of the
\pkg{casebase} package is to provide practitioners with an easy-to-use
software tool to predict the risk (or cumulative incidence) of an event,
conditional on a particular patient's covariate profile. Our package
retains the flexibility of case-base sampling and the familiar interface
of the \code{glm} function.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to discretize the study base into
an infinite amount of \emph{person moments}. These person moments are
indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure status
and an outcome status attached to it. We note that there is only a
finite number of person moments associated with the event of interest
(what Hanley \& Miettinen call the \emph{case series}). The case-base
sampling refers to the sampling from the base of a representative finite
sample called the \emph{base series}.

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras; the interested
reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
Saarela \citeyearpar{saarela2016case}. First, let
\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
event of interest for individual \(i=1, \ldots,n\). For simplicity, we
will consider Type I censoring due to the end of follow-up at time
\(\tau\) (the general case of non-informative censoring is treated in
Saarela \citeyearpar{saarela2016case}). We assume a continuous time
model, which implies that the counting process jumps are less than or
equal to one. We are interested in modeling the hazard functions
\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]

@@

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual contributions of the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating equation for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modeling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, this sampling strategy corresponds
essentially to a time-homogeneous Poisson process with hazard equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
in the base series, and \(B\) is the total population-time for the study
base. More complex examples are also available; see for example Saarela
\& Arjas \citeyearpar{saarela2015non}, where the probabilities of the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. With this sampling mechanism,
Saarela \& Arjas are able to increase the efficiency of their
estimators, when compared to uniform sampling.

@@

The instantaneous failure rate is independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

For competing-risk analyses with \(J\) possible events, we can show that
each person-moment's contribution of the likelihood is of the form

@@

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is a vast ecosystem of \code{R}
packages implementing different methods. In this section, we describe
how the functionalities of \pkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Several \proglang{R} packages implement methodologies for competing risk
analysis; for a different perspective, see Mahani \& Sharabiani
\citeyearpar{mahani2015bayesian}. The package \pkg{cmprsk} provides
methods for cause-specific subdistributions, such as in the Fine-Gray
model \citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific cumulative incidence functions from
unadjusted, non-parametric survival functions. Our package
\pkg{casebase} also provides functionalities for competing risk analysis
by estimating the cause-specific hazards, and from these quantities, we
can derive the cause-specific cumulative incidence functions.

Finally, several packages include functions to estimate the cumulative
incidence function. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explictly modeled. Using this estimate, the survival function and the
cumulative incidence function can be obtained using their functional
relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
this approach for both single-event and competing-event analyses. The
second category outlined above broadly corresponds to Cox models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator). Packages that implement this approach
include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the cumulative incidence function, whereas the second category produces
estimates in the form of step-wise functions. This was one of the main
motivations for introducting case-base sampling in survival analysis.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  case-base sampling;
\item
  estimation of the smooth hazard function;
\item
  calculation of the risk function.
\end{enumerate}

By separating the sampling and estimation functions, we allowed the
possibility of users implementing more complex sampling scheme, as
described in Saarela \citeyearpar{saarela2016case}.

@@

The fitting function \code{fitSmoothHazard} starts by looking at the
class of the dataset: if it was generated from \code{sampleCaseBase}, it
automatically inherited the class \code{cbData}. If the dataset supplied
to \code{fitSmoothHazard} does not inherit from \code{cbData}, then the
fitting function starts by calling \code{sampleCaseBase} to generate the
base series. In other words, users can bypass \code{sampleCaseBase}
altogether and only worry about the fitting function
\code{fitSmoothHazard}.

@@

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using Lasso
  or elastic-net. This functionality is provided through the
  \code{glmnet} package \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \code{gam} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \code{gbm} package. This feature is still experimental.
\end{itemize}

@@

We created \code{absoluteRisk} as an \code{S3} generic, with methods
that match the different types of outputs of \code{fitSmoothHazard}. The
method dispatch system of \proglang{R} then takes care of matching the
correct output to the correct methodology for calculating the cumulative
incidence function, without requiring any intervention from the user.

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}

@@

\begin{CodeInput}
R> c(
R+   "Exp." = AIC(casebase_exponential),
R+   "Weib." = AIC(casebase_time),
R+   "Splines" = AIC(casebase_splines)
R+ )
\end{CodeInput}

@@

\begin{CodeInput}
R> 
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
R+   const(Phase) + const(Source) + const(Age),
R+ data = bmtcrr, cause = 1, model = "fg"
R+ )
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-3support-data}{%
\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next case study, we explore an
advanced feature.

For the third case study, we will demonstrate regularization during
hazard estimation. Since we are using logistic regression, existing
packages, like \pkg{glmnet} can be leveraged. To demonstrate this
feature, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within 5 hospitals in the
United States of America. Imputation was conducted using the \pkg{mice}
package in \proglang{R} with default settings \citep{mice}. The imputed
dataset is available as part of the \pkg{casebase} package. For more
information please see the SUPPORT data documentation in the package.

we have a total of 9104 observations and 30 variables, two of which are
SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology
score (aps). These scores were developed for a period of 180 days, which
makes our comparison over 5.56 years exploratory. Therefore, we will
assess these two models before using either as our baseline, by
comparing them to a full model using all variables. Also, our single
imputation introduces bias in the data. As the goal here is purely
demonstrative, this bias will be consistent with any of the following
models. As such, their comparison is still valid, though any
interpretation requires an awareness of said bias. A description of
covariates and response variables can be found in Table
\ref{tab:support1} and a breakdown of each categorical variable in Table
\ref{tab:support2}. The original data was obtained from the Department
of Biostatistics at Vanderbilt University \citep{harrell_2020}.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

Before generating an absolute risk curve using regularized logistic
regression, we will explore three options as a baseline comparison. The
original study compared these scores over a span of 180 days, which may
not be comparable over the full 5.56 years. Therefore, We will examine
three cox regression scenarios: sps only, aps only and all covariates
excluding sps and aps. Next, we will compare each modeling using the
Brier score \citep{graf1999ass}. The Brier score can demonstrate both
discrimination and calibration in a single model \citep{graf1999ass} and
provide a comparable metric between semi-parametric and parametric
methods in the survival context. For all the following analyses in this
case study, we will use 80\% of the observations to train our models,
while 20\% will be used to generate absolute risk curves, all of which
will be averaged to demonstrate each model's predictive ability. Over
5.56 years, \ref{fig:baselines} demonstrates that all the cox models
have similar, if not identical average predictions. Figure
\ref{fig:brier1} examines the brier scores as they change over time. As
the full model has the lowest brier score on overall, it will be used in
the following absolute risk comparison to regularized cox, regularized
logistic regression and the unadjusted Kaplan-Meier curve.

{\centering \includegraphics{../figures/baselines-1}

\caption{\label{fig:baselines} Absolute risk curves estimated from hazard models. Red: APS only hazard was used to generate the absolute risk. Green: All covariates aside from APS and SPS. Blue: SPS only. Purple: unadjusted Kaplan-Meier curve.}\label{fig:baselines}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:brier1} Brier scores over survival time. Full model (Full) performs best, while the other two (APS only and SPS only) are comparable. Red: APS only model was used to generate the brier scores. Green: all covariates excluding APS and SPS. Blue: SPS only model.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

To get the absolute risk following regularized cox regression, we had to
work around limitations in existing packages. To make use of the
\pkg{survival} package, we first estimate the hazard coefficients using
thecoxnet implementation for lasso penalization in \pkg{glmnet}
\citep{regpathcox}. Next, we fit a cox model using the survival package
with the selected variables only, from our regularized fit. Finally, we
replaced the coefficients in the survival object with those from our
glmnet fit, allowing us to make use of the absolute risk tools available
in the \pkg{survival} package.

@@

\begin{CodeInput}
R> # hazard for cox using glmnet
R> coxNet <- glmnet::cv.glmnet(
R>   x = x, y = y, family = "cox", alpha = 1,
R>   standardize = TRUE
R> )
R> # getting all non-zero covariates
R> nonzero_covariate_cox <- predict(coxNet,
R>   type = "nonzero",
R>   s = "lambda.1se"
R> )
R> # keeping coefficients for later
R> nonzero_coef_cox <- coef(coxNet, s = "lambda.1se")
R> # cutting dataset to only include nonzero covariates
R> cleanCoxData <- as.data.frame(cbind(y, x[, nonzero_covariate_cox]))
R> # create a skeleton object
R> coxNet <- survival::coxph(Surv(time = d.time, event = death) ~ .,
R>   data = cleanCoxData, x = TRUE
R> )
R> # replace coefficients with regularized ones
R> coxNet$coefficients <- nonzero_coef_cox@x
\end{CodeInput}
\end{CodeChunk}

For our penalized logistic regression model, we opted for the natural
log of time, which assumes a Weibull distribution for our event. As
casebase makes use of the glmnet package, we interact with the
fitsmoothhazard.fit function using a matrix interface, where \emph{y}
contains the time and event variables and \emph{x} contains all
covariates we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables. Here,
we used lasso penalization by setting alpha to 1, leading to our smooth
hazard function.

>>>>

From the perspective of clinicians and their patients, the most relevant
quantity is often the 5- or 10-year risk of having a certain event given
the patient's particular circumstances, and not the hazard ratio between
a treatment and control group. Therefore, to make sound clinical
decisions, it is important to accurately estimate the \emph{full} hazard
function, which can then be used to estimate the cumulative incidence
function (CIF). Since the parametric hazard is a smooth function of
time, the CIF and the survival function estimates also vary smoothly
over time.

@@

In this article, we present an \proglang{R} package that combines the
ideas of Hanley \& Miettinen into a simple interface. The purpose of the
\pkg{casebase} package is to provide practitioners with an easy-to-use
software tool to predict the risk (or cumulative incidence) of an event,
conditional on a particular patient's covariate profile. Our package
retains the flexibility of case-base sampling and the familiar interface
of the \code{glm} function. In addition, we provide additional features,
including performance metrics and visualization tools.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to discretize the study base into
an infinite amount of \emph{person moments}. These person moments are
indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure status
and an outcome status attached to it. We note that there is only a
finite number of person moments associated with the event of interest
(what Hanley \& Miettinen call the \emph{case series}). Case-base
sampling refers to the sampling from the base of a representative finite
sample called the \emph{base series}.

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras; the interested
reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
Saarela \citeyearpar{saarela2016case}. First, let
\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
event of interest for individual \(i=1, \ldots,n\). For simplicity, we
will consider Type I censoring due to the end of follow-up at time
\(\tau\) (the general case of non-informative censoring is treated in
Saarela \citeyearpar{saarela2016case}). We assume a continuous time
model, which implies that the counting process jumps are less than or
equal to one. We are interested in modelling the hazard functions
\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]

@@

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual contributions of the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating equation for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modelling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

In Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the authors
suggest sampling the base series \emph{uniformly} from the study base.
In terms of Poisson processes, their sampling strategy corresponds
essentially to a time-homogeneous Poisson process with hazard equal to
\(\rho_i(t) = b/B\), where \(b\) is the number of sampled observations
in the base series, and \(B\) is the total population-time for the study
base (e.g.~the sum of all individual follow-up times). More complex
examples are also possible; see for example Saarela \& Arjas
\citeyearpar{saarela2015non}, where the intensity functions for the
sampling mechanism are proportional to the cardiovascular disease event
rate given by the Framingham score. With this sampling mechanism,
Saarela \& Arjas are able to increase the efficiency of their
estimators, when compared to uniform sampling.

@@

In this model, the instantaneous failure rate is independent of
\(t\).\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Case-base sampling can also be used in the context of competing-risk
analyses. Assuming there are \(J\) competing events, we can show that
each person-moment's contribution of the likelihood is of the form

@@

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is already a vast ecosystem of \code{R}
packages implementing different methodologies. In this section, we
describe how the functionalities of \pkg{casebase} compare to these
packages.

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our review to packages that implement at least
one of the following features: parametric modeling, non-proportional
hazard models, competing risk analysis, penalized estimation, and CIF
estimation. By searching for appropriate keywords in the
\code{DESCRIPTION} file of these packages, we found 60 relevant
packages. These 60 packages were then manually examined to determine
which ones are comparable to \pkg{casebase}. In particular, we excluded
packages that were focused on a different set of problems, such as
frailty and multi-state models. The remaining 14 packages appear in
Table \ref{tab:surv-pkgs}, along with some of the functionalities they
offer.

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}
\citeyearpar{survival-package}. The types of models they allow vary for
each package. For example, \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas both \pkg{flexsurv}
and \pkg{survival} allow users to supply any distribution of their
choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and
\pkg{rstpm2} also have the ability to model the effect of time using
splines, which allows flexible modeling of the hazard function.
Moreover, \pkg{flexsurv} has the ability to estimate both scale and
shape parameters for a variety of parametric families (see Table
\ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear
combination of covariates (including time). Therefore, our package
allows the user to model the effect of time using splines. Also, by
including interaction terms between covariates and time, it also allows
users to fit time-varying coefficient models. However, we do not
explicitly model any shape parameter, unlike \pkg{flexsurv}.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation in the context of parametric hazards. To our
knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to
offer this functionality.

Next, several \proglang{R} packages implement methodologies for
competing risk analysis; for a different perspective on this topic, see
Mahani \& Sharabiani \citeyearpar{mahani2015bayesian}. The package
\pkg{cmprsk} provides methods for cause-specific subdistributions, such
as in the Fine-Gray model \citeyearpar{fine1999proportional}. On the
other hand, the package \pkg{CFC} estimates cause-specific CIFs from
unadjusted, non-parametric survival functions. Our package
\pkg{casebase} also provides functionalities for competing risk analysis
by estimating parametrically the cause-specific hazards. From these
quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explictly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of step-wise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducting case-base sampling in survival analysis.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  case-base sampling;
\item
  estimation of the smooth hazard function;
\item
  estimation of the CIF.
\end{enumerate}

By separating the sampling and estimation functions, we allow the
possibility of users implementing more complex sampling scheme (as
described in Saarela \citeyearpar{saarela2016case}), or more complex
study designs (e.g.~time-varying exposure).

@@

Next, the fitting function \code{fitSmoothHazard} starts by looking at
the class of the dataset: if it was generated from
\code{sampleCaseBase}, it automatically inherited the class
\code{cbData}. If the dataset supplied to \code{fitSmoothHazard} does
not inherit from \code{cbData}, then the fitting function starts by
calling \code{sampleCaseBase} to generate the base series. In other
words, users can bypass \code{sampleCaseBase} altogether and only worry
about the fitting function \code{fitSmoothHazard}.

@@

\begin{itemize}
\tightlist
\item
  \code{glm}: This is the familiar logistic regression.
\item
  \code{glmnet}: This option allows for variable selection using Lasso
  or elastic-net (cf.~Case Study 3). This functionality is provided
  through the \code{glmnet} package \citep{friedman2010jss}.
\item
  \code{gam}: This option provides support for \emph{Generalized
  Additive Models} via the \code{gam} package
  \citep{hastie1987generalized}.
\item
  \code{gbm}: This option provides support for \emph{Gradient Boosted
  Trees} via the \code{gbm} package. This feature is still experimental.
\end{itemize}

@@

@@

\begin{CodeInput}
R> plot(pt_object_strat, facet.params = list(ncol = 2))
\end{CodeInput}

@@

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R+   "Weib." = AIC(casebase_time),
R+   "Splines" = AIC(casebase_splines))
\end{CodeInput}

@@

\begin{CodeInput}
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
R+                         const(Phase) + const(Source) + const(Age),
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-3support-data}{%
\section{Case study 3---SUPPORT Data}\label{case-study-3support-data}}

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. For the third case study, we show how
\pkg{casebase} can also be used for variable selection through
regularized estimation of the hazard function. To do this, we simply
replace logistic regression with a penalized counterpart. In
\pkg{casebase}, we achieve this by using the \pkg{glmnet} package
\citep{friedman2010jss}.

To illustrate this functionality, we use the dataset from the Study to
Understand Prognoses Preferences Outcomes and Risks of Treatment
(SUPPORT) \citep{knaus1995support}. The SUPPORT dataset tracks death in
five American hospitals within individuals who are considered seriously
ill.~The original data is available online from the Department of
Biostatistics at Vanderbilt University \citep{harrell_2020}. The cleaned
and imputed data consists of 9104 observations and 30 variables, and it
is available as part of the \pkg{casebase} package. For more information
about this dataset, the reader is encouraged to look at the
documentation in our package. A description of the variables can be
found in Table \ref{tab:support1}, and a breakdown of each categorical
variable appears in Table \ref{tab:support2}.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
\label{tab:support2}
\end{table}

In the original study, the authors developed two scores for predicting
the outcome. These scores are available as part of the data: SUPPORT day
3 physiology score (\code{sps}) and APACHE III day 3 physiology score
(\code{aps}). Using these scores, we will explore three Cox regression
scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only,
\code{aps} only, and all covariates excluding \code{sps} and \code{aps}.
We will compare the three models using a time-dependent version of the
classical Brier score that is adjusted for censoring
\citep{graf1999ass}. The Brier score can be used to assess both
discrimination and calibration in a single model. Moreover, it provides
a metric that can be used to compare parametric and semi-parametric
survival models. To account for overfitting, we will use 95\% of the
observations to train our models, while the remaining 5\% will be used
to compute absolute risk curves and the Brier scores.

\begin{CodeInput}
R> data(support)
R> # Change time to years
R> support$d.time <- support$d.time/365.25
R> 
R> # Split into test and train
R> train_index <- sample(nrow(support), 0.95*nrow(support))
R> test_index <- setdiff(1:nrow(support), train_index)
R> 
R> train <- support[train_index,]
R> test <- support[test_index,]
\end{CodeInput}
\end{CodeChunk}

In Figure\ref{fig:brier1} (A), we can see that all Cox models have
similar average survival curves. On the other hand, Figure
\ref{fig:brier1} (B) shows that the full model has the lowest Brier
score overall. For this reason, we will use this model in our comparison
between regularized Cox regression, Kaplan-Meier estimation, and the
approach from \pkg{casebase}.

@@

\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

For our penalized logistic regression model, we opt for the natural log
of time; recall that this corresponds to a Weibull distribution. For
fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which
is a matrix interface to the function \code{fitSmoothHazard}. In order
to use this interfact, we first create a matrix \code{y} containing the
time and event variables, and a matrix \code{x} containing all remaining
covariates we would like to include in the model space. We used lasso
penalization by setting \code{alpha = 1}. Moreover, we use the argument
\code{penalty.factor} to force the time variable into the final model.

@@

\begin{CodeInput}
R> x <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = train)[, -c(1)] # Remove intercept
R> y <- data.matrix(subset(train, select = c(d.time, death)))
R> 
R> # Regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
R+   family = "glmnet",
R+   time = "d.time", event = "death",
R+   formula_time = ~ log(d.time), alpha = 1,
R+   ratio = 10, standardize = TRUE,
R+   penalty.factor = c(0, rep(1, ncol(x)))
R+ )
\end{CodeInput}
\end{CodeChunk}

Next, we can use the object \code{cbFull} to estimate the absolute risk
curve on the test data. We first create a matrix \code{newx} containing
the relevant variables. Next, using \code{times}, we specify the time
points at which the cumulative incidence will be computed. Finally,
using the parameter \code{s = "lambda.1se"}, we specify at which value
of the tuning parameter we want to retrieve the coefficient estimates.
